Corrected Figures

### 2021-12-15 corrected_figures.Rmd
results corrected

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs and behaviors in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on 183 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

```{r setup, include=FALSE}
# Before the first commit, I added "here", changed the load paths with here(), and added the %!in% function

knitr::opts_chunk$set(echo = FALSE)
library(ggdag)
library(ggrepel)
library(here)
library(tidyverse)
'%!in%' <- function(x,y)!('%in%'(x,y))

# Introduction

Culture is an important part of social life but cultures are continuously evolving. In 1972, for example, more than 40 percent of US adults supported a law outlawing interracial marriage. Three decades later, this opinion had become so uncommon that the question was removed from the US General Social Survey. How does this kind of cultural change happen?

Attempts to account for opinion changes in society have produced conflicting theories about the process of opinion formation and the ability of individuals to maintain consistent attitudes. Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues. As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden]. By contrast, cohort replacement theories posit that people do hold opinions and are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder]. Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p. xi; @achen1992]. More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-declarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].

In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 183 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that large numbers of U.S. adults changed their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses. At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to persist than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes and behaviors consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to their general political beliefs. We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

# Theoretical Background

## Belief Formation in Cultural Sociology

How does cultural change happen at the individual level? Let us make the question concrete by imagining a person who answers the same question each year for several years. The question could be anything, but assume it is this GSS question: "Please tell me whether you strongly agree, agree, disagree, or strongly disagree [with this statement]: 'a working mother can establish just as warm and secure a relationship with her children as a mother who does not work.'"

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions. We will explain this in greater detail below. For now, consider the following two simple models:

\begin{equation}
  y_{it} = y_{it-1} + \nu_{it}
  (\#eq:AR)
\end{equation}

\begin{equation}
  y_{it} = U_i + \nu_{it}
  (\#eq:FE)
\end{equation}

These models may seem similar at first glance, but they have different implications for the pattern of individuals' responses we would observe over time. Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below. Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences. In the next two sections, we consider each of these models and briefly discuss their connections to sociological theories.

[Figure \@ref(fig:dag) about here]

```{r}
library(ggdag)
library(dagitty)
```

## Active Updating Model

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin]. In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.

More informally, this AUM posits a person who is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors. When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model. When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time. Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.

Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities [@gross09; @degloma]. As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.

The AUM makes no assumption about the distribution of $\nu_{it}$. Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$. This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.

Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller]. In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.

## Settled Dispositions Model

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are thus temporary influences only. At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp. 66ff]. In other words, updating happened, but it happened in the past, prior to the time of the study. In this sense, $U_i$ reflects the "habitus."

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder]. Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.

On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case. The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse]. Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse]. As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.

Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level. For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported. Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].

The two general models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time. In the SDM, changes tend to revert because each person returns to their baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.

## Consistency in Response

Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items. As such, we designate a second term, _consistency_, to reflect this second dimension of change. _Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.

## Toward Theoretical Synthesis

We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating. Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.

The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern. This disagreement thus forms the primary structure of our analysis. However, it is possible that some items display active updating while others do not. There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.

_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman]. To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein]. This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space. Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442].

The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition. Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency. On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman]. In other words, liberal Republicans should become liberal Democrats, not conservative Republicans. This means that partisan identification should display high levels of updating. Again, this does not mean that many people change, only that changes tend to persist.

Alongside ideological identification, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman]. For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy. Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently. There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.

Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs. However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse]. However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.

_Public vs. Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances. However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time. However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we expect to find evidence of active updating throughout the population, not just among young people.

Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.

As we noted previously, we do not expect that all items will align with just one model. If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions. Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work. We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

# Research Questions and Expectations

With these considerations in mind, we ask the following questions:

First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18). However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.

Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors. As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.

# Analytic Strategy

To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below. While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.

## Statistical Model

### Basic Models

Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time. We first consider measuring the amount of active updating. Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:

\begin{equation}
  E(y_{i3}) = y_{i2}
  (\#eq:AUM2)
\end{equation}

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2. If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.

The SDM makes the following prediction:

\begin{equation}
  E(y_{i3}) = U_i
  (\#eq:SDM2)
\end{equation}

Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:

\begin{equation}
  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
  (\#eq:SDM2B)
\end{equation}

That is, the SDM predicts that the average response of previous waves is the best predictor of the next response. If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.

### Combined Model

Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor. If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition. Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$. If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model. However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating. We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:

\begin{equation}
  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
  (\#eq:MODEL)
\end{equation}

Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B). If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.

Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating. We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model. If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$. If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.

### Comparison to Other Approaches

Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test. Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS responses. This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error. These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier. While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves. This approach also combines persisting and non-persisting change into two similar but distinct kinds of change: "structural" and "non-structural," both of which can be persisting and non-persisting. This distinction, while important for some theoretical questions, is not our focus.

A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief. This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_. Since we at times invoke both these models, we include in our analysis several composite scales of related items. If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in their levels of active updating.

Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal]. While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns. One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster. If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 would become a better predictor of wave 3.

We do not view this as a problem for our theoretical models. If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant. This might lead to a higher estimate of active updating than we might observe in a population that did not take the survey, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update beliefs over time.

A second form of panel conditioning posits that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time, even if they actually change. In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change. This would be problematic for our study, as it would under-estimate the amount of real change in the population. However, if this is the case, we should observe no active updating in responses and high levels of consistency, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.

Finally, a number of approaches exist for evaluating theoretical process of belief formation and change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization]. These tools are well designed to address the questions they set out to answer. However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.

### Limitations of the Method

Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach and therefore limit the conclusions we can reach. The first is that our model is designed to allocate variance explained to each of the prior waves rather than to assign probabilities to each data-generating process. Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.

The second is measurement error, which is a form of non-persisting change. For some researchers, measurement error represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure. For other theories, measurement error reflects individuals' inability to accurately report their response. It is also possible that measurement error reflects errors of selection and interpretation, such as misunderstanding the question or incorrect coding.

Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), $\phi$ estimates will be biased toward .5, since responses with error would be random departures from baseline. There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings]. On the other hand, previous studies of reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them. Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.

The third challenge of our analysis is that we focus on predicting wave 3. If individuals have a high likelihood of changing between waves 2 and 3, our ability to predict responses at wave 3 will be limited and $\beta$ will be low. Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3. If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error, and it may not be reasonable to consider this sort of change "persistent."

In addition to these three challenges, there are two forms of change that our model is not well designed to account for. The first is a uni-directional shock to the population. Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates. The second is change in the variance of responses. If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will be absorbed into $\beta$ but not enter into $\phi$.

Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error. Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model. We can only detect whether there is any evidence of persisting belief changes, and therefore whether there is any evidence that active updating is taking place in the population.

## Analysis Steps

Our analysis proceeds in three steps to answer our three research questions. First, we evaluate the overall evidence in support of the active updating model. To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$. We calculate the posterior probability that the model with the free parameter fits the data better. If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.[^link]

[^link]: All data used in this analysis is publicly available through the General Social Survey at the National Opinion Research Center at http://gss.norc.org/. R code for how we cleaned the GSS panels, all analyses performed in this paper, and all figures created in this paper can be found at http://github.com/krkiley/panel_change.

Second, for variables that show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is concentrated among younger respondents. To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff. Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45. We test a range of cutoffs to ensure robustness of the overall pattern to specific ages.

Finally we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories. Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be inductive.

## Item Selection

To test our model on as broad a range of items as possible, we sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels. We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer. We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on subject material. Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about government spending, and questions about social life.

We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust. This includes a six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany"). We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are. We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman]. Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy. We combine four parallel questions about how frequently individuals socialize to create a "social life" scale. We combine four questions about support for suicide under different circumstances. We also created a scale of support for police use of violence against criminal suspects by combining five binary questions about the conditions under which individuals support police use of violence.

In total, we test the model on 183 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves. Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.

[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options. These results are reported in Appendix B.

# Results

```{r, message = FALSE, warn = FALSE}
load(here("data", "results_df.Rdata"))
load(here("data", "attitude_vars.Rdata"))

@@

Our model estimates two parameters of interest for each GSS item: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3. High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating. Our measure of active updating, $\phi$, captures the relative proportion of wave 3 variance predicted by wave 2. If responses are generated through a true settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3). As the evidence of active updating increases, $\phi$ will increase toward 1. Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3.

## Evidence for Active Updating

To evaluate the evidence in favor of the active updating model, we compare for all 183 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$. If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.

>>>>

---
title: "Corrected Figures"
author: "Kevin Kiley"
date: "12/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggdag)
library(ggrepel)
library(here)
library(tidyverse)
'%!in%' <- function(x,y)!('%in%'(x,y))
```

```{r, message = FALSE, warn = FALSE}
load(here("correction", "results_df.Rdata"))
load(here("data", "attitude_vars.Rdata"))

@@

```{r loadfiguredata}
load(here("correction", "figure_data.Rdata"))
```

```{r loadcoarse}
#load(here("correction", "coarse_vars.Rdata"))
